TacticToe: Learning to Prove with Tactics

### 2018-03-19 tactictoe_jv.tex
Plan and diagram in the introduction

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and prettification of the generated proof  facilitates user
interaction.
\end{itemize}

\todoi{graphic explaining how each part relate to each other}

@@

The different components depicted in Fig~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}. 
  At the basis of those predictions are 
  similarity measures computed from syntactic features of mathematical formulas 
  in Section~\ref{sec:features}.
  Modification of the knowledge base performed by orthogonalization 
  in Section~\ref{sec:ortho} and abstraction in Section~\ref{sec:synthesis}
  improves the quality of the databases.
  And the preselection method in Section~\ref{sec:presel} increases the speed 
  of the prediction algorithm.
  The learned predictors relying on this knowledge serves as
  guidance for the proof search algorithm in Section~\ref{sec:proofsearch} 
  which explores different proof paths in the proof tree in 
  Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  And generated tactic-level proofs are analysed in 
  Section~\ref{sec:case_study}.

@@

\section{Proof Presentation}\label{sec:proofdisplay}

@@

\paragraph{Prettification}
Without prettification, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
in a specific state, we can strip unnecessary information, such as
module prefixes. If possible, we group all local declarations in the
script under a single \texttt{let} binding at the start of the script. We
also replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

The total effect of these minimization and prettification techniques
on the readability of a proof script is depicted in
Example~\ref{ex:pretty}.

\begin{example}(Theorem \texttt{EVERY\_MAP})\label{ex:pretty}\\\
Proof script before minimization and prettification:
\begin{lstlisting}[language=SMLSmall]
boolLib.REWRITE_TAC [DB.fetch "list" "EVERY_CONJ", DB.fetch "list" "EVERY_MEM",
  DB.fetch "list" "EVERY_EL", ldots ,  combinTheory.o_DEF] THEN
BasicProvers.Induct_on [HolKernel.QUOTE " (*#loc 1 11380*)l"] THENL
  [BasicProvers.SRW_TAC [] [],
   simpLib.ASM_SIMP_TAC (BasicProvers.srw_ss ()) [boolLib.DISJ_IMP_THM,
     DB.fetch "list" "MAP", DB.fetch "list" "CONS_11", boolLib.FORALL_AND_THM]]

\end{lstlisting}
Proof script after minimization and prettification:
\begin{lstlisting}[language=SMLSmall]
Induct_on `l` THENL
  [SRW_TAC [] [], ASM_SIMP_TAC (srw_ss ()) [DISJ_IMP_THM, FORALL_AND_THM]]
\end{lstlisting}
\end{example}

>>>>

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and embellishment of the generated proof  facilitates user
interaction.
\end{itemize}

@@

The different components depicted in Fig~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}. 
  At the basis of those predictions are 
  similarity measures computed in Section~\ref{sec:features} from syntactic 
  features of mathematical formulas .
  Modification of the knowledge base performed by orthogonalization 
  in Section~\ref{sec:ortho} and abstraction in Section~\ref{sec:synthesis}
  improves the quality of the databases.
  And the preselection method described in Section~\ref{sec:presel} increases 
  the speed 
  of the prediction algorithm.
  The learned predictors relying on this knowledge serves as
  guidance for the proof search algorithm presented in 
  Section~\ref{sec:proofsearch} 
  which explores different proof paths in the proof tree defined in 
  Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  And generated tactic-level proofs are analysed in 
  Section~\ref{sec:case_study}.

@@

\section{Minimization and Embellishment}\label{sec:proofdisplay}

@@

\paragraph{Embellishment}
Without embellishment, the returned tactic is barely readable as it contains
information to guarantees that each \sml subterm is interpreted in the way in
any context. Since we return the proof at a point where the \sml interpreter is
in a specific state, we can strip unnecessary information, such as
module prefixes. If possible, we group all local declarations in the
script under a single \texttt{let} binding at the start of the script. We
also replace extended terms by their quoted version.
All in all, if a prettified tactic $t_p$ has the same effect its original
$t_o$, we replace $t_p$ by $t_o$ in the proof.\\

The total effect of these minimization and embellishment techniques
on the readability of a proof script is depicted in
Example~\ref{ex:pretty}.

\begin{example}(Theorem \texttt{EVERY\_MAP})\label{ex:pretty}\\\
Before minimization and embellishment:
\begin{lstlisting}[language=SMLSmall]
boolLib.REWRITE_TAC [DB.fetch "list" "EVERY_CONJ", DB.fetch "list" "EVERY_MEM",
  DB.fetch "list" "EVERY_EL", ldots ,  combinTheory.o_DEF] THEN
BasicProvers.Induct_on [HolKernel.QUOTE " (*#loc 1 11380*)l"] THENL
  [BasicProvers.SRW_TAC [] [],
   simpLib.ASM_SIMP_TAC (BasicProvers.srw_ss ()) [boolLib.DISJ_IMP_THM,
     DB.fetch "list" "MAP", DB.fetch "list" "CONS_11", boolLib.FORALL_AND_THM]]

\end{lstlisting}
After minimization and embellishment:
\begin{lstlisting}[language=SMLSmall]
Induct_on `l` THENL
  [SRW_TAC [] [], ASM_SIMP_TAC (srw_ss ()) [DISJ_IMP_THM, FORALL_AND_THM]]
\end{lstlisting}
\end{example}
